---
title: "PS2_NSW"
author: "1093122"
format: html
editor: visual
---

## Exercise 1

### a.

```{r}
library(tidyverse)
library(haven)
library(dplyr)
library(broom)

experimental <- read_dta("https://users.nber.org/~rdehejia/data/nsw_dw.dta")

```

### b.

```{r}
experimental <- experimental |> 
  rename(earnings74 = re74,
         earnings75 = re75,
         earnings78 = re78)
experimental
```

### c.

```{r}
experimental <- experimental |> 
  mutate(race = case_when(black == 1 ~ "black",
                          hispanic == 1 ~ "hispanic",
                          TRUE ~ "white")
         ) |> 
  select(-black, -hispanic)
experimental
```

### d.

```{r}
experimental <- experimental |> 
  mutate(treat = ifelse(treat == 1, "treated", "non_treated"),
         degree = ifelse(nodegree == 1, "no_high_school", "high_school"),
         marriage = ifelse(married, "married", "unmarried")
         ) |> 
  select(-nodegree, -married)
experimental
```

### e.

```{r}
experimental <- experimental |> 
  mutate(employment74 = ifelse(earnings74==0, "unemployed", "employed"),
         employement75 = ifelse(earnings75==0, "unemployed", "employed"))

experimental
```

### f.

```{r}
experimental <- experimental |> 
  select(-data_id)
experimental
```

```{r}
# define the cleanup function for data cleaning later
cleanup <- function(experimental){
  experimental <- experimental |> 
  rename(earnings74 = re74,
         earnings75 = re75,
         earnings78 = re78)
  experimental <- experimental |> 
  mutate(race = case_when(black == 1 ~ "black",
                          hispanic == 1 ~ "hispanic",
                          TRUE ~ "white")
         ) |> 
  select(-black, -hispanic)
  experimental <- experimental |> 
  mutate(treat = ifelse(treat == 1, "treated", "non_treated"),
         degree = ifelse(nodegree == 1, "no_high_school", "high_school"),
         marriage = ifelse(married, "married", "unmarried")
         ) |> 
  select(-nodegree, -married)
  experimental <- experimental |> 
  mutate(employment74 = ifelse(earnings74==0, "unemployed", "employed"),
         employement75 = ifelse(earnings75==0, "unemployed", "employed"))
  experimental <- experimental |> 
  select(-data_id)
}
```

## Exercise 2

### a.

```{r}
library(modelsummary)
datasummary_skim(experimental, type = "numeric")
datasummary_skim(experimental, type = "categorical")
```

### b.

The two groups are similar in terms of their average age, education and earnings in 1974. The average earnings for the treated group are slightly higher in 1975 and much higher in 1978 than the untreated group. The race composition in treated and untreated groups are similar while there is a slightly higher proportion of hispanic people and a lower proportion of white people in the non-treated group. A higher proportion of people in the treated group have high school degreee compared with the non-treated group. The two groups also have similar marriage patterns and employment status in 1974. In 1975, a higher proportion of people from the treated group are employed. There seems to be no obvious selection bias based on observable characteristics.

```{r}
datasummary_balance(experimental
                    ~ treat, data = experimental)
```

### c.

The entire 95% confidence interval lies above 0. This indicates the NSW program significantly increased participants' earnings.

```{r}
# calculate the group means for the treated and non-treated group
group_means <- experimental |> 
  group_by(treat) |> 
  summarize(
    mean_earnings78 = mean(earnings78),
    se = sd(earnings78)/sqrt(n()),
    n = n()
  )

# Compute ATE
ate <- group_means$mean_earnings78[2] - group_means$mean_earnings78[1]

# Pooled standard error
se_ate <- sqrt(sum(group_means$se^2))

# Construct confidence interval
ci_lower <- ate - 1.96 * se_ate
ci_higher <- ate - 1.96 * se_ate
ci <- c(ci_lower = ci_lower, ci_higher = ci_higher)
ci
```

## Exercise 3

### a.

```{r}
cps_controls <- read_dta("https://users.nber.org/~rdehejia/data/cps_controls.dta")
cps_controls
```

### b.

```{r}
cps_controls <- cleanup(cps_controls)
cps_controls
```

### c.

```{r}
experimental_treated <- experimental |> 
  filter(treat == "treated")
composite <- bind_rows(cps_controls, experimental_treated)
```

### d.

```{r}
balance_table <- datasummary_balance(
  ~ treat,  # Compare groups
  data = composite,
  title = "Balance Table: NSW-treated vs CPS-controls",
  dinm = FALSE
)

# Print the table
balance_table
```

### e.

The CPS controls appear to be very different from the treated group for all the observable characteristics listed above. The raw differences in earnings78 can not be interpreted as the average treatment effect as there is likely to be a selection bias and other characteristics such as race and education are likely to confound the treatment effect estimate.

## Exercise 4

### a.

The estimated treatment effect is \$1066 when regressing earnings on treatment status and covariates in the composite dataset with a p-value of 0.054. This suggest weak evidence of a positive effect. However, the average treatment effect obtained from the randomised experiment is larger (\$1794) and statistically significant at 5% significant level as shown in 2.c.

```{r}
reg1 <- lm(earnings78~., composite)
summary(reg1)
```

### b.

The assumption of conditional independence is needed. After controlling for observable covariates, treatment assignment must be independent of potential outcomes. This requires that all confounders are observed and included in the model.

### c.

The ATE conditional on covariates is expressed as:

$$
\text{ATE} = \mathbb{E}[Y \mid D = 1, X] - \mathbb{E}[Y \mid D = 0, X]
$$

Under **unconfoundedness**, the assignment of the treatment can be seen as random. Hence, D is mean independent of $(Y_0, Y_1)$, we can write ATE over the full populatioon as:

$$
\text{ATE} = \mathbb{E}[\mathbb{E}[Y_1 \mid X] - \mathbb{E}[Y_0 \mid X]]
$$

Using the linear conditional expectations above:

$$
\mathbb{E}[Y_1 \mid X] = \alpha_1 + X' \beta_1 \\
\mathbb{E}[Y_0 \mid X] = \alpha_0 + X' \beta_0
$$

So,

$$
\mathbb{E}[Y_1 \mid X] - \mathbb{E}[Y_0 \mid X] = (\alpha_1 - \alpha_0) + X'(\beta_1 - \beta_0)
$$

Now take the **expectation over the distribution of** ( X ):

$$
\text{ATE} = (\alpha_1 - \alpha_0) + \mathbb{E}[X'](\beta_1 - \beta_0)
$$

### d.

We are given that the conditional expectation of the potential outcomes is:

$$
\mathbb{E}[Y_d \mid X] = \alpha_d + X' \beta_d \quad \text{for } d = 0,1
$$

Using the switching equation:

$$
Y = D \cdot Y_1 + (1 - D) \cdot Y_0,
$$

we can compute:

$$
\mathbb{E}[Y \mid D, X] = D \cdot (\alpha_1 + X' \beta_1) + (1 - D)(\alpha_0 + X' \beta_0)
$$

Expanding this, we get:

$$
\mathbb{E}[Y \mid D, X] = \alpha_0 + X' \beta_0 + D \cdot \left[(\alpha_1 - \alpha_0) + X'(\beta_1 - \beta_0)\right]
$$ Therefore, we should run a regression to estimate ATE as follows:

$$
Y = \alpha + \tau D + X' \gamma + \varepsilon
$$ Additional assumptions:\
(1) It is a linear model.\
(2) Unconfoundedness: $(Y_0, Y_1) \perp\!\!\!\perp D \mid X$

## Exercise 5

### a.

In the propensity score model, we should exclude earnings78 from the list of covariates because it is a post-treatment outcome variable. Including it would introduce post-treatment bias, violating the assumption that the propensity score is based solely on pre-treatment characteristics. Only covariates determined before treatment assignment should be included when estimating the propensity score.

```{r}
logit_data <- composite |> 
  mutate(
    treat = ifelse(treat == "treated", 1, 0),
    
    # Convert categorical vars to factor first
    race = as.factor(race),
    degree = as.factor(degree),
    marriage = as.factor(marriage),
    employment74 = as.factor(employment74),
    employement75 = as.factor(employement75)
  ) 

reg2 <- glm(treat~.-1 - earnings78, data = logit_data, family = binomial(link = "logit"))
summary(reg2)
```

```{r}
logit_data <- augment(reg2, type.predict = "response") |> 
  rename(propensity_score = .fitted)
```

### b.

```{r}
treated <- logit_data |> 
  filter(treat == 1)

ggplot(treated, aes(x = propensity_score)) +
  geom_histogram(position = "identity", bins = 30) +
  labs(
    title = "Propensity Score Distributions for the treated group",
    x = "Propensity Score",
  ) +
  theme_minimal()
```

```{r}
control <- logit_data |> 
  filter(treat == 0)

ggplot(control, aes(x = propensity_score)) +
  geom_histogram(position = "identity", bins = 30) +
  labs(
    title = "Propensity Score Distributions for the control group",
    x = "Propensity Score",
  ) +
  theme_minimal()
```

```{r}
# extra summary statistics
logit_data %>%
  group_by(treat) %>%
  summarise(
    mean_ps = mean(propensity_score),
    sd_ps = sd(propensity_score),
    min_ps = min(propensity_score),
    max_ps = max(propensity_score)
  )

```

The result highlights the substantial difference in the distribution of propensity scores for the control and treated group. The histogram of the control group’s propensity scores shows an extreme concentration near 0, while the treated group’s scores are widely distributed between 0 and 0.65. This means people in the treated group are more likely to receive treatment than the control group, which suggests substantial differences in observed covariates between groups.

### c.

```{r}
# Compute IPW weights
logit_data <- logit_data |> 
  mutate(
    weight_treat = treat / propensity_score,
    weight_control = (1 - treat) / (1 - propensity_score)
  )

# Compute psw
psw_estimate <- mean(
  logit_data$weight_treat * logit_data$earnings78 -
  logit_data$weight_control * logit_data$earnings78
)

psw_estimate

```

### d.

```{r}
# Step 1: Trim observations with extreme propensity scores
logit_trimmed <- logit_data %>%
  filter(propensity_score >= 0.1, propensity_score <= 0.9)

# Step 2: Compute psw
logit_trimmed <- logit_trimmed %>%
  mutate(
    weight_treat = treat / propensity_score,
    weight_control = (1 - treat) / (1 - propensity_score)
  )

# Step 3: Compute trimmed psw estimator
psw_estimator_trimmed <- mean(
  logit_trimmed$weight_treat * logit_trimmed$earnings78 -
  logit_trimmed$weight_control * logit_trimmed$earnings78
)

psw_estimator_trimmed

```

### e.

The difference between parts (c) and (d) arises from the presence of extreme propensity scores in the untrimmed sample. In part (c), these lead to very large inverse weights, especially among control units with scores near 0. This results in a highly volatile and unreliable ATE estimate. In contrast, part (d) trims these problematic observations and ensures that only observations with sufficient overlap are used for causal inference, where treated and control units are more comparable. This produces a much more stable and credible estimate of the treatment effect.
