---
title: "PS1_FREDR"
author: "1093122"
format: html
editor: visual
---

## Exercise 1: Getting familiar with the FRED API

Install the fredr package and load the library; Set API key

```{r, message=FALSE}
library(tidyverse)
library(fredr)
library(tibble)
my_key <- Sys.getenv("api_key")
fredr_set_key(my_key)
```

get access to GDP quarterly data from 1947

```{r}
fredr(series_id = "GDP")
```

Modify the API call to retrieve quarterly GDP data from January 01, 2000, and save it in a tibble called gdp.

```{r}
gdp <- fredr(series_id = "GDP", 
             observation_start = as.Date("2000-01-01"))
```

Find out what other time series fredr has

```{r}
fredr_category_children(category_id = 0)
```

Retrieve unemployment rates

```{r}
fredr_series_search_text("unemployment")
```

```{r}
u <- fredr(series_id = "UNRATE", observation_start = as.Date("2000-01-01"))
u
```

## Exercise 2: Plotting macroeconomic patterns

### a. Plot the monthly unemployment rates stored in u using ggplot2

```{r}
u |> 
  ggplot(aes(x = date, y = value)) +
  geom_line() +
  scale_x_date(
  breaks = seq(as.Date("2000-01-01"), as.Date("2025-01-01"), by = "2 years"),
  date_labels = "%Y"
)+
  xlab("Date") +
  ylab("Percent")+
  labs(title = "Unemployment Rate (%)", 
       subtitle = "Frequency: Monthly; Time period: 2000-01 to 2025-04; Seasonally Adjusted", 
       caption = "Data Source: U.S. Bureau of Labor Statistics via FRED®") +
  theme_bw()
```

### b. Description of unemployment rates change during the business cycles.

Unemployment rates increase significantly during recessions. In 2001, it increased from around 4% to 6%. In 2008, it increased from 5% to around 10%. During the Covid-19 recession in 2020, it increased sharply from 3% to around 15%.Unemployment then decreases during economic recoveries and economic booms, as shown by the downward trend after each recession.

### c. GDP Plot

```{r}
gdp |> 
  ggplot(aes(x = date, y = value)) +
  geom_line() +
  scale_x_date(
  breaks = seq(as.Date("2000-01-01"), as.Date("2025-01-01"), by = "2 years"),
  date_labels = "%Y"
)+
  xlab("Date") +
  ylab("Billions of Dollars") +
  labs(title = "Gross Domestic Product", 
       subtitle = 
      "Frequency: Quarterly; Time period: 2000-01 to 2025-04; Seasonally Adjusted Annual Rate", 
       caption = "Data Source: U.S. Bureau of Economic Analysis via FRED®") +
  theme_bw()

```

### d. Create quarter-over-quarter growth and annualised QoQ for plotting GDP growth

```{r}
gdp <- gdp |>
  mutate(qoq = ((value - lag(value))/lag(value)),
        qoq_annualised = (1 + qoq )^4 - 1, # as if the qoq growth continues for a year, what would be the resulting annual growth rate
        qoq = qoq*100,
        qoq_annualised = qoq_annualised*100)
```

We do not have a previous period for the first observation of this dataset (2000-01-01). Therefore, for the first period, we do not obtain a quarter-over-quarter growth and the values are missing.

### e. Plot both measures of quarterly growth in the same plot

```{r}
gdp |> 
  ggplot(aes(x = date)) +
  geom_line(aes(y = qoq, color = "qoq")) +
  geom_line(aes(y = qoq_annualised, color = "qoq_annualised"))+
  scale_x_date(
  breaks = seq(as.Date("2000-01-01"), as.Date("2025-01-01"), by = "2 years"),
  date_labels = "%Y"
)+
  xlab("Date") +
  ylab("Percent")+
  labs(title = "Gross Domestic Product Growth", 
       subtitle = "Frequency: Quarterly; Time period: 2000-04 to 2025-04", 
       caption = "Data Source: U.S. Bureau of Economic Analysis via FRED®") +
  scale_color_manual(name = "Growth Metrics",
                     values = c("qoq" = "#1f77b4",
                      "qoq_annualised" = "#ff7f0e"),
                     labels = c("qoq" = "QoQ (%)",
                                "qoq_annualised" = "Annualised QoQ(%)"))+
  theme_bw()
```

### f. Pull QoQ growth from FRED API directly and fix the missing values

```{r}
gdp_growth <- fredr(series_id = "GDP", units = "pch", observation_start = as.Date("2000-01-01"))
gdp_growth
```

The quarterly growths rates match to my qoq variable.

```{r}
gdp$qoq[1] <- gdp_growth$value[1]
gdp$qoq_annualised[1] <- ((1+gdp$qoq[1]*0.01)^4 - 1)*100
gdp
```

## Exercise 3: Spotting recessions

```{r}
recessions <- tibble(
  start = as.Date(c("2001-03-01", "2007-12-01", "2020-02-01")),
  end   = as.Date(c("2001-11-30", "2009-06-30", "2020-04-30"))
)
```

### a. Plot unemployment rates and add shading for recession periods

```{r}
uplot <-  u |> 
  ggplot(aes(x = date, y = value)) +
   geom_rect(
    data = recessions,
    aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
    fill = "grey",
    inherit.aes = FALSE) +
  geom_line() +
    scale_x_date(
  breaks = seq(as.Date("2000-01-01"), 
               as.Date("2025-01-01"), 
               by = "5 years"),
  date_labels = "%Y"
)+
  xlab("Date") +
  ylab("Percent")+
  theme_bw()
uplot
```

### b. Plot QoQ GDP growth with recession shading using geom_rect()

```{r}
gdp_growth_plot <- gdp_growth |> 
  ggplot(aes(x = date, y = value)) +
   geom_rect(
    data = recessions,
    aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf), # Inf / -Inf to cover the whole column
    fill = "grey", # use grey color
    inherit.aes = FALSE) +
  geom_line() +
  scale_x_date(
  breaks = seq(as.Date("2000-01-01"), 
               as.Date("2025-01-01"), by = "5 years"),
  date_labels = "%Y"
)+
  xlab("Date") +
  ylab("Percent")+
  theme_bw()
gdp_growth_plot
```

### c. Combine the plots next to each other

```{r}
library(patchwork)
uplot + gdp_growth_plot+
  plot_annotation(title = "Unemployment (Left, %) and GDP Growth (Right, %)", 
                  caption = "Data Source: FREDR and NBER")
```
